RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning

### 2019-09-30 docs/paper/paper.tex
docs: Revise abstract and introduction

\begin{abstract}
The notion of creativity is wider than generating something inspired by reference pieces of art. It also includes generation of something that meets criteria of being an art regardless of similar pieces existence. However, supervised machine learning is limited by the former scope, because it requires a dataset. Reinforcement learning can be used instead if the goal is to reveal more aspects of creativity. In this paper, a cross-entropy agent is trained to compose musical pieces by interacting with a piano roll environment. The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory.
\end{abstract}

@@

Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation. To name a few of common notations, there are sheet music, tablature, and MIDI standard. It is not required from output representation to unambiguously define sound waveform. For example, sheet music leaves exact loudness of played notes to discretion of performer and may include only imprecise hints like pianissimo (very quiet). Anyway, there are parameters of sounds that must be determined by their representation. Usually, such parameters are pitch, start time, and duration.

@@

Plenty of various approaches for composing music automatically exist. In particular, there are approaches based on machine learning. Both supervised learning and reinforcement learning are applicable here. Moreover, some researchers combine them \cite{jaques2016generating}.

In this paper, a pure reinforcement learning approach to music generation is studied. The reasons for not involving supervised learning at all are as follows:
\begin{itemize}
	\item Finding new ways of music creation is a more challenging task than imitation of famous pieces. If no known pieces are used, chances are that the harder problem is considered and it is not replaced with the simpler problem of imitation.
	\item There are tuning systems other than equal temperament (for instance, in microtonal music). For some of them it may be impossible to collect dataset large enough to allow training models in a supervised fashion. However, developers of a tuning system should know some underlying principles and so (at least, in theory) it is possible to create evaluational rules and train an agent based on them.
\end{itemize}

Results reported at this draft version are far from using above advantages at full scale. The current study is rather a proof-of-concept for composition of music with reinforcement learning solely.

@@

\section{Methodology}
\label{sec:approach}

@@

\section{Further Improvements}
\label{sec:improvements}
% start from supervisedly pre-trained weigths;

>>>>

\begin{abstract}
The notion of creativity is wider than generating something inspired by reference pieces of art. It also includes generation of something that meets criteria of being an art, but is not backed by existing pieces. On the one hand, supervised machine learning is limited by the former scope, because it requires a dataset. Reinforcement learning can be applied instead if the goal is to reveal more aspects of creativity. On the other hand, numerous recent advances in music generation are achieved with supervised deep learning. In this paper, a method that uses both reinforcement learning and deep learning is developed for automated music composition. An agent with neural network as actor model is trained to compose musical pieces by interacting with a piano roll environment. The environment scores submitted pieces based on some hand-written evaluational rules derived from music theory. No dataset is involved, because cross-entropy method is used instead of gradient-based methods.
\end{abstract}

@@

Algorithmic music composition is automatic generation of outputs representing musical pieces and written in some formal notation. To name a few of common notations, there are sheet music, tablature, and MIDI standard. It is not required from output representation to unambiguously define sound waveform. For example, sheet music leaves exact loudness of played notes to discretion of a performer and may include only imprecise hints like pianissimo (very quiet). Anyway, there are parameters of sounds that must be determined by their representation. Usually, such parameters are pitch, start time, and duration.

@@

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer. Thus, there is an open research problem. Some recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating}.

In this paper, a new approach to music composition is suggested. There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent can be either playing a note at the current time step or moving to the next time step. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method.

Obviously, the approach belongs to reinforcement learning, because there is an environment, but a dataset is absent. Nevertheless, there are strong connections with supervised deep learning. From generation of new pieces point of view, actor model returns probabilities of next notes given current state. This is similar to sampling new sequences from a model trained with maximum likelihood method on a dataset of existing sequences.

More details on the methodology are provided in Section \ref{sec:methodology}, but here it is appropriate to discuss its advantages.

The reasons for not involving supervised training at all are as follows:
\begin{itemize}
	\item Finding new ways of music creation is a more challenging task than imitation of famous pieces. If no known pieces are used, chances are that the harder problem is considered and it is not replaced with the simpler problem of imitation.
	\item There are tuning systems other than equal temperament (for instance, in microtonal music). For some of them it may be impossible to collect dataset large enough to allow training models in a supervised fashion. However, developers of a tuning system should know some underlying principles and so (at least, in theory) it is possible to create evaluational rules and train an agent based on them.
\end{itemize}

Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested. The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or other methods (genetic algorithms, ant colony algorithms, evolutionary strategies \cite{salimans2017evolution}).

Actually, results reported at this draft version are far from using above advantages at full scale. The current study is rather a proof-of-concept, but this proof-of-concept is easily extendable and ideas on how it can be improved are listed in Section \ref{sec:improvements}. Also it is worth mentioning that the suggested approach is quite straightforward, but to the best of the author's knowledge it is not described anywhere.

@@

\section{Methodology}
\label{sec:methodology}
% dataset -> rules; backpropagation -> cross-entropy; learning rate -> weights std.

@@

\section{Further Improvements}
\label{sec:improvements}
% better evaluational rules;
% randomized initial observations;
% evolutionary strategies instead of cross-entropy;
% start from supervisedly pre-trained weigths;
